---
title: "Probabilistic alternative to 2 group t-test"
author: "Rasmus Hindström"
date: last-modified
format:
    html:
        toc: true
        toc_float: true
        embed-resources: true
---

# 0. Summary

This report demonstrates the use and interpretation of a probabilistic alternative
to the two sample t-test. For a more expansive deep dive into the topic refer to
Matti Vuorre's blog post [How to Compare Two Groups with Robust Bayesian Estimation in R](https://vuorre.com/posts/2017-01-02-how-to-compare-two-groups-with-robust-bayesian-estimation-using-r-stan-and-brms/).
Much of which is adapted within the context of microbiome research. 

A common task in microbiome research is to compare alpha diversity between
two or more groups. Often done with the t-test
and concluded as a binary "significant or non-significant" result, the more
rich inference provided by the probabilistic comparison will become apparent.
Alpha diversity is a measure of the within sample diversity, and it is often computed
as the number of observed species (richness), the Shannon index, or the Simpson index. 
Here we demostrate comparisons of the Shannon index which takes into account abundance
information, and is typically close to normally distributed.  

# 1. Comparison of two groups

Here a dataset from the `mia` package is used to demonstrate the comparison and 
interpretation differences between the two sample t-test (Welch t-test) and an
alternative probabilistic approach with Bayesian estimation using the `brms` package.

```{r}
#| label: load-libraries
#| code-summary: Load required libraries
#| echo: false
#| output: false 

library(mia)
library(dplyr)
library(brms)
library(bayesplot)
```

```{r}
#| label: prepare-data
#| code-summary: prepare example data
#| 
data("peerj13075", package = "mia")
tse <- peerj13075

# Add Shannon index
tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = "shannon"
)

# Construct data frame
df <- as.data.frame(colData(tse))
```

## 1.1. Welch t-test

Welch t-test makes frees the assumption of equal variances between the two groups.
Still retaining the assumption of normality.

In essance the t-test compares the means of two groups, and returns a p-value
indicating the probability of observing the data if the null hypothesis is true.
The null hypothesis states that there is no difference between the two groups.

```{r}
#| label: welch-t-test
#| code-summary: Perform Welch t-test

t.test(shannon ~ Diet, data = df, var.equal = FALSE)
```

The p-value suggests that there is no significant difference between the two groups.
We fail to reject the null hypothesis.

## 1.2. Bayesian estimation

Approaching the same problem with Bayesian estimation using the `brms` package.

As Matti Vuorre (2017) points out, the t-test is only a linear model under the hood.
Extending this idea, we can formulate a probabilistic model and use `brms` to estimate
the parameters of interest. In this case the group means and variances.
Using a Student's t-distribution and estimating $\nu$ allows for a more robust model
that is less sensitive to outliers.

$$y_{ik} \sim \mathcal{T}(\nu, \mu_k, \sigma_k^2)$$

$$ \mu_k = \beta_0 + \beta_k$$

$$ \sigma_k^2 = \gamma_0 + \gamma_k$$

```{r}
#| label: bayesian-estimation
#| code-summary: Perform Bayesian estimation

# Fit Bayesian model (default priors)
fit <- brm(
    formula = bf(
        shannon ~ Diet,
        sigma ~ Diet
    ),
    data = df,
    family = student,
    iter = 4000,
    chains = 4,
    cores = 4
)
```

```{r}
#| label: model-summary
  
summary(fit)
```

A straight forward interpretation of the model can be made from the 95% credible
intervals (CIs). Here our baseline group is the Mixed diet. The intercept for 
DietVeg shows the difference in the Shannon index compared to the Mixed diet. 

The CI for the intercept of DietVeg indicates that the difference between the 
two groups is likely to not be meaningful, as the CI includes 0. 

## 1.3. Model diagnostics

```{r}
#| label: model-diagnostics
#| code-summary: Diagnostics plots 
plot(fit)
pp_check(fit)
```

The model summary shows non-pathogenic Rhat values, indicating that the chains have converged.
Effective sample sizes are sufficiently high (> 1000) for each parameter.
Chains are well mixed in the traceplots.
The posterior predictive plot shows that the model is able to capture the observed data well enough.
However, the two peaked nature of the data is not captured.



# Bibliography

- Kruschke, J. K. (2013). Bayesian estimation supersedes the t test. Journal of Experimental Psychology: General, 142(2), 573–603. https://doi.org/10.1037/a0029146

- Vuorre, Matti. 2017. “How to Compare Two Groups with Robust Bayesian Estimation in R.” January 2, 2017. [https://vuorre.com/posts/2017-01-02-how-to-compare-two-groups-with-robust-bayesian-estimation-using-r-stan-and-brms/](https://vuorre.com/posts/2017-01-02-how-to-compare-two-groups-with-robust-bayesian-estimation-using-r-stan-and-brms/).








