---
title: "Probabilistic multi-group comparison of alpha diversity"
author: "Rasmus Hindström"
date: last-modified
format:
    gfm:
        toc: true
---

# 0. Summary

This report demonstrates the use and interpretation of a probabilisitc alternative
to multi-group comparisons of alpha diversity. The basis is a multi-level model
with a group-level effect, which is estimated using Bayesian estimation with the `brms` package.


# 1. Data preparation

Preparing data from the `mia` package.

```{r}
#| label: load-libraries
#| code-summary: Load required libraries
#| echo: true
#| output: false
library(mia)
library(dplyr)
library(brms)
library(bayesplot)
```

```{r}
#| label: prepare-data
#| code-summary: prepare example data
data("peerj13075", package = "mia")
tse <- peerj13075
tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = "shannon"
)
df <- as.data.frame(colData(tse))
```

# 2. Model fitting

The model is fitted using the `brm` function from the `brms` package.
Reponse variable is the Shannon index, and the grouping variable is the 3 classes
of `age`; `Adult`, `Middle_age`, and `Elderly`.

Model definition is as follows:

$$
y_{ik} \sim \text{t}(\nu, \mu_{ik}, \sigma_k)
$$

$$
\mu_{ik} = \beta_0 + \beta_k
$$

$$
\sigma_k = \gamma + \gamma_k
$$

*Default priors used by `brm()`*

$$
\nu \sim \gamma(2, 0.1)
$$
$$
\beta_0 \sim \text{t}(3, 1.3, 2.5)
$$
$$
\beta_k, \sigma \sim \text{t}(3, 0, 2.5)
$$


```{r}
#| label: fit-model
#| code-summary: Fit Bayesian model
#| warning: false
#| output: false 

# Model with partial pooling
fit <- brm(
    formula = bf(
        shannon ~ 1 +(1|Age),
        sigma ~ 1 + (1|Age)
    ),
    data = df,
    family = student(),
    iter = 8000,
    chains = 6,
    cores = 6
) 
```

```{r}
#| label: fit-summary
#| echo: false
#| output: true
    
summary(fit)

coefficients(fit)
```

```{r}
#| label: plotting-post
#| code-summary: Posterior plotting
#| code-fold: true
#| fig-height: 6
#| fig-width: 12  
  
library(ggplot2)
library(patchwork)

draws <- as_draws_df(fit)
plot_data <- data.frame(
    population = draws$b_Intercept,
    pop_mean = mean(draws$b_Intercept),
    adult = draws$b_Intercept + draws$'r_Age[Adult,Intercept]',
    elderly = draws$b_Intercept + draws$'r_Age[Elderly,Intercept]',
    middle_age = draws$b_Intercept + draws$'r_Age[Middle_age,Intercept]'
)

p1 <- ggplot(data = plot_data) +
    geom_density(aes(x = adult), fill = "blue", alpha = 0.5, color = "blue") +
    geom_density(aes(x = elderly), fill = "purple", alpha = 0.6, color = "purple") +
    geom_density(aes(x = middle_age), fill = "orange", alpha = 0.7, color = "orange") +
    geom_vline(xintercept = plot_data$pop_mean, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = "Posterior Distributions of Group means",
        x = "Shannon index",
        y = "Density"
    ) +
    annotate(
        "text", x = 2, y = 2.5,
        label = "Orange: Middle Aged\nPurple: Elderly\nBlue: Adult",
    ) +
    theme_minimal()

p2 <- ggplot(data = plot_data) +
    geom_boxplot(aes(y = adult, x = 1), fill = "blue", alpha = 0.7, color = "black") +
    geom_boxplot(aes(y = elderly, x = 2), fill = "purple", alpha = 0.7, color = "black") +
    geom_boxplot(aes(y = middle_age, x = 3), fill = "orange", alpha = 0.7, color = "black") +
    geom_hline(yintercept = plot_data$pop_mean, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(
        breaks = c(1, 2, 3),
        labels = c("Adult", "Elderly", "Middle Age")
    ) +
    labs(
        title = "Boxplots of Group Means",
        x = "Group",
        y = "Shannon Index"
    ) +
    theme_minimal()

p1 + p2
```

From the plots we can infer groups are not similar. Particularly the 
Middle aged (Orange) group appears to have a lower Shannon index. The boxplots 
paint a clear picture of the higher overlap between the Adult and Edlerly group,
while the Middle aged group differs. In both plots the red dashed line indicates 
the total population posterior mean.

# 3. Classical approach

With classical methods multi group comparisons are often done in multiple steps. 

```{r}
#| label: kruskal-test

kruskal.test(shannon ~ Age, df)
```

We get a p-value < 0.05, so there are 'significant' differences within the groups.

```{r}
pairwise.wilcox.test(df$shannon, df$Age, p.adjust = "fdr")
```

Post-hoc testing requires adjusting the p-values. Here we observe
that the groups `Middle_age` and `Adult` have a significant difference.
Other pairwise comparisons are not significant. 

# 4. Conclusions

Both the probabilistic and classical approach to multigroup testing gives the same result. 
The middle age group is signaled out. Contrasting the two methods, it is clear
that the probabilistic approach gives a richer inference, with the added benefit
of intuitive interpretation. 

Note: there are some warnings from the sampler of the probabilistic model,
but Rhats are close to 1 and ESS are not too low, if we disregard the estimates for 
the sigma term. Some tweaks to the model definition might be required to get cleaner
sampling. 